---
title: "RaggedExperiment"
output:
  BiocStyle::html_document:
    toc: true
vignette: |
  %\VignetteIndexEntry{RaggedExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `r Biocpkg("RaggedExperiment")` package provides a flexible data
representation for copy number, mutation and other ragged array schema for
genomic location data. It aims to provide a framework for a set of samples
that have differing numbers of genomic ranges.

The `RaggedExperiment` class takes a `GRangesList` representation and provides
a semblance of a rectangular dataset. Dimensions of the `RaggedExperiment`
correspond to the number of ranges in the entire dataset and the number
of samples represented in the data.

# Constructing a `RaggedExperiment` object

```{r, echo = FALSE}
suppressPackageStartupMessages({
    library(RaggedExperiment)
    library(GenomicRanges)
})
```

```{r, eval = FALSE}
library(RaggedExperiment)
```

We start with a couple of `GRanges` objects each representing an individual
sample: 

```{r}
sample1 <- GRanges(
    c(GENEA = "chr1:1-10:-", GENEB = "chr1:11-18:+"),
    score = 3:4)
sample2 <- GRanges(
    c(GENEC = "chr2:1-10:-", GENED = "chr2:11-18:+"),
    score = 1:2)
``` 

Include column data `colData` to describe the samples:

```{r}
colDat <- DataFrame(id = 1:2)
```

## Using `GRanges` objects

```{r}
ragexp <- RaggedExperiment(sample1 = sample1,
                           sample2 = sample2,
                           colData = colDat)
```

Access the dimension names

```{r}
dimnames(ragexp)
```

## Using a `GRangesList` instance

```{r}
grl <- GRangesList(sample1 = sample1, sample2 = sample2)
RaggedExperiment(grl, colData = colDat)
```

# *Assay methods

It provides several different functions for representing ranged data in
a rectangular matrix via the `*Assay` methods.

## sparseAssay
The most straightforward matrix representation of a `RaggedExperiment` will
return a matrix of dimensions equal to the product of the number of ranges and
samples.

```{r}
sparseAssay(ragexp)
```

