---
title: "RaggedExperiment"
author:
- name: Martin Morgan
  affiliation: Roswell Park Cancer Institute, Buffalo, NY
- name: Marcel Ramos
  affiliation: Roswell Park Cancer Institute, Buffalo, NY
date: "`r BiocStyle::doc_date()`"
vignette: |
  %\VignetteIndexEntry{RaggedExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document2:
    toc_float: true
Package: RaggedExperiment
---

```{r, echo = FALSE}
suppressPackageStartupMessages({
    library(RaggedExperiment)
    library(GenomicRanges)
})
```

# Introduction

The `r Biocpkg("RaggedExperiment")` package provides a flexible data
representation for copy number, mutation and other ragged array schema for
genomic location data. It aims to provide a framework for a set of samples
that have differing numbers of genomic ranges.

The `RaggedExperiment` class derives from a `GRangesList` representation and
provides a semblance of a rectangular dataset. The row and column dimensions of
the `RaggedExperiment` correspond to the number of ranges in the entire dataset
and the number of samples represented in the data, respectively.

# Installation

```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
BiocInstaller::biocLite("RaggedExperiment")
```

Loading the package:

```{r, eval = FALSE}
library(RaggedExperiment)
```

# Constructing a `RaggedExperiment` object

We start with a couple of `GRanges` objects, each representing an individual
sample: 

```{r}
sample1 <- GRanges(
    c(GENEA = "chr1:1-10:-", GENEB = "chr2:15-18:+"),
    score = 3:4)
sample2 <- GRanges(
    c(GENEC = "chr1:1-10:-", GENED = "chr2:11-18:+"),
    score = 1:2)
``` 

Include column data `colData` to describe the samples:

```{r}
colDat <- DataFrame(id = 1:2)
```

## Using `GRanges` objects

```{r}
ragexp <- RaggedExperiment(sample1 = sample1,
                           sample2 = sample2,
                           colData = colDat)
ragexp
```

## Using a `GRangesList` instance

```{r}
grl <- GRangesList(sample1 = sample1, sample2 = sample2)
RaggedExperiment(grl, colData = colDat)
```

## Using a `list` of `GRanges`

```{r}
rangeList <- list(sample1 = sample1, sample2 = sample2)
RaggedExperiment(rangeList, colData = colDat)
```

## Using a `List` of `GRanges` with metadata

__Note__: In cases where a `SimpleGenomicRangesList` is provided along with
accompanying metadata (accessed by `mcols`), the metadata is used as the
`colData` for the `RaggedExperiment`. 

```{r}
grList <- List(sample1 = sample1, sample2 = sample2)
mcols(grList) <- colDat
RaggedExperiment(grList)
```

# Accessors

## Range data

```{r}
rowRanges(ragexp)
```

## Dimension names

```{r}
dimnames(ragexp)
```

## `colData`

```{r}
colData(ragexp)
```

# Subsetting

## by dimension

Subsetting a `RaggedExperiment` is akin to subsetting a `matrix` object. Rows
correspond to genomic ranges and columns to samples or specimen. It is possible
to subset using `integer`, `character`, and `logical` indices.

## by genomic ranges

The `overlapsAny` and `subsetByOverlaps` functionalities are available for use
for `RaggedExperiment`. Please see the corresponding documentation in
`RaggedExperiment` and `GenomicRanges`.

# *Assay functions

`RaggedExperiment` package provides several different functions for representing
ranged data in a rectangular matrix via the `*Assay` methods.

## sparseAssay

The most straightforward matrix representation of a `RaggedExperiment` will
return a matrix of dimensions equal to the product of the number of ranges and
samples.

```{r}
dim(ragexp)
Reduce(`*`, dim(ragexp))
sparseAssay(ragexp)
length(sparseAssay(ragexp))
```

## compactAssay

Samples with identical ranges are placed in the same row. Non-disjoint ranges
are **not** collapsed. 

```{r}
compactAssay(ragexp)
```

## disjoinAssay

This function returns a matrix of disjoint ranges across all samples. Elements
of the matrix are summarized by applying the `simplify` functional argument
to assay values of overlapping ranges. 

```{r}
disjoinAssay(ragexp, simplify = mean)
```

## qreduceAssay

The `qreduceAssay` function works with a `query` parameter that highlights
a window of ranges for the resulting matrix. The returned matrix will have
dimensions `length(query)` by `ncol(x)`. Elements contain assay values for the
_i_ th query range and the _j_ th sample, summarized according to the `simplify`
functional argument.

First we define our summary function that calculates a weighted average score
per query range. _Note_ that there are three arguments to this function. Please
see the documentation `?qreduceAssay` for more details.

```{r}
weightedmean <- function(scores, ranges, qranges)
    sum(scores * width(ranges)) / sum(width(ranges))
```

A call to `qreduceAssay` involves the `RaggedExperiment`, the `GRanges` query
and the `simplify` functional argument. 

```{r}
qreduceAssay(ragexp,
             query = GRanges(c("chr1:1-10:-", "chr2:11-18:+")), 
             simplify = weightedmean)
```

# Coercion

The `RaggedExperiment` provides a family of parallel functions for coercing to
the `SummarizedExperiment` class. By selecting a particular assay index (`i`), a
parallel assay coercion method can be achieved. 

Here is the list of function names: 

* `sparseSummarizedExperiment`
* `compactSummarizedExperiment`
* `disjoinSummarizedExperiment`
* `qreduceSummarizedExperiment`

See the documentation for details.

# Session Information

```{r}
sessionInfo()
```

